 

### 2021-06-22 vertical-thermal-review.Rmd
misc edits

#SI figure numbers
SIfigs_NEON_all = "S1-S#"

#SI figure captions

@@

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(3) leaf temperature ($T_{leaf}$);
(2) the leaf traits that most strongly influence leaf temperature and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including scaling across space and time.*

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest itself.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; `r fig_NEON_vertical`; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height (Parker 1995).

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**. Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
 The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density (LAD) ie. leaf area per unit volume, increases with tree-height eventually forming the canopy layer.

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In this review, we focus on growing season conditions unless otherwise noted.

@@

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. Noteably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

**Leaf thermoregulation is controlled by leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016]. Therefore, leaf traits are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**

Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004] where degrees of direct to diffused light and resultant water demands shape sun vs. shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`). At greater tree height, both biological variables (tree-height) and biophysical variables (light, temperature, and evaporative demand )(Fig. `r fig_NEON_vertical`) also play a role in shaping leaf traits.

@@

>>>>

#SI figure numbers
SIfig_NEON_all = "S1"

#SI figure captions
SIfig_example_legend <- "Figure S1. Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)"

@@

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsWithincanopyVariationRate2004a].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence leaf temperature and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including scaling across space and time.*

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest itself.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Figs. `r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height (Parker 1995).

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**. Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest. 
Leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests. 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

@@

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. Noteably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

**Leaf thermoregulation is controlled by leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016]. Therefore, leaf traits are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**

Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004] where degrees of direct to diffused light and resultant water demands shape sun vs. shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`). At greater tree height, both biological variables (tree-height) and biophysical variables (light, temperature, and evaporative demand )(Fig. `r fig_NEON_vertical`) also play a role in shaping leaf traits.

@@

SIfig_example_legend

### 2021-06-22 vertical-thermal-review_SI.Rmd
misc edits

#SI figure numbers
SIfig_example = "S1"

#SI figure captions
SIfig_example_legend <- "Figure S1. ...."

@@

# Appendix `r appendix_SCBI_treering`. `r appendix_SCBI_treering_title`

\newpage
## References

>>>>

#SI figure numbers
SIfig_NEON_all = "S1"

#SI figure captions
SIfig_example_legend <- "Figure S1. Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)"

@@

# `r SIfig_example_legend`
![**Figure `r SIfig_NEON_all`. Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)**. Shown are  height profiles in July mean ± 1 standard deviation for maximum photosyntehtically active ratiation (PAR), maximum wind speed, minimum humidity, maximum $T_{air}$, and maximum biological temperature, $T_{bio}$. Sites are grouped into the following categories: (A) (sub)subtropical and warm temperate broadleaf: ..., (B) temperate open/ savanna forests: ..., (C) temperate mesic broadleaf forests: ..., (D) temperate conifer forests: ..., (E) northern and boreal forests: ...  [Issue #35](https://github.com/EcoClimLab/vertical-thermal-review/issues/35). ](NEON_height_profiles/figures/profile_all_groups.png)

## References
